|

### 2019-08-15 article/article.Rmd

---
title:  |
        | Bottom-Up Accountability and Infrastructure Delivery: Mixed Evidence from Brazil
author: 
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
- "Jessica Voigt[^voigt]"
date: \today
thanks: "We thank Fábio Barros, Guilherme Jardim Duarte, Stephen Herzog, Catarina Roman and David Skarbek for their helpful comments. We are specially grateful to the staff of Transparência Brasil for their excellent research assistance. This research was pre-registered on the EGAP pre-registry tool (<https://egap.org/registration/4505>) and approved by the Brazilian Ministry of Education. The authors thank the 2016 Google Social Impact Challenge for financial support and declare there are no conflict of interest. Replication data and code are available at <http://github.com/umbertomignozzetti/tdp-accountability>."
abstract: "We study the effect of a mobile phone application designed to increase citizens' ability to monitor public works in Brazil. The app allows individuals to find delayed school constructions in 1030 Brazilian cities and send anonymous messages to politicians requesting information about construction delays and completion times. Our results show the app has a null impact on the completion times subsequently reported by the municipalities. Additionally, we find that few politicians react to citizens' requests. These findings suggest it is difficult to motivate bottom-up accountability in new democracies, especially when politicians are unresponsive to non-electoral pressures. This paper has implications for the design of bottom-up accountability mechanisms."
abstractspacing: double
keywords: accountability, Brazil, field experiment, impact evaluation, state capacity, technology
fontsize: 12pt
margin: 
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
---

@@

We use the TDP app to conduct two experimental interventions and test its impact on six outcomes related to school completion rates and complaints to public authorities. Overall, our results show that providing information to citizens has very limited impact on policy outcomes. We find that the TDP app increased the likelihood of school completion by 6.8 percent in the first experiment, but the effect is null in the second intervention. We also find that the app increased the percentage of the executed budget by 157.5 percent in our second experiment, but only after nearest neighbourhood matching adjustment. No result reported in the first intervention or in the unmatched sample. Importantly, both interventions have no significant impact on any of the other measures of school investment and completion rates we employed in this study.

@@

# Load packages
pkgs <- c('tidyverse', 'dummies', 'kableExtra',
          'haven', 'compareGroups', 
          'devtools', 'lubridate', 'brazilmaps',
          'abjutils', 'lfe', 'cobalt', 
          'MatchIt', 'ri2', 'DeclareDesign', 'knitr',
          'DiagrammeR', 'DiagrammeRsvg', 'rsvg', 
          'flextable', 'officer')
invisible(lapply(pkgs, require, character.only = T))

@@

```{r model1, echo=FALSE, warning=FALSE, message=FALSE, results='hide', cache=TRUE}
load('tdp.RData')

# Load packages
pkgs <- c('tidyverse', 'dummies', 'kableExtra',
          'haven', 'compareGroups', 
          'devtools', 'lubridate', 'brazilmaps',
          'abjutils', 'lfe', 'cobalt', 
          'MatchIt', 'ri2', 'DeclareDesign', 'knitr',
          'DiagrammeR', 'DiagrammeRsvg', 'rsvg', 
          'flextable', 'officer')
invisible(lapply(pkgs, require, character.only = T))

@@

resInt1 <- calcATEs(outcs, treats, ctrls, fes, cls, dat = impEvalph1)
```

```{r results1, results='asis', echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE}
# Results intervention 1
library(stargazer)
stargazer(resInt1[[2]][[4]],resInt1[[2]][[8]],resInt1[[2]][[12]],
          resInt1[[2]][[16]],resInt1[[2]][[20]],resInt1[[2]][[24]],
          title = 'Impact Evaluation -- Intervention 1', 
          keep='treatTreatment', type = 'latex',
          covariate.labels = 'ATE',
          column.labels = c('Investment\\\\ Before','Invest\\\\ After',
                             'Delta\\\\ Investment','Finished\\\\ Construction',
                             'Cancelled\\\\ Construction','Updated\\\\ Date'),
          dep.var.labels.include=F,
          omit.stat=c('adj.rsq', 'ser'), style='apsr',
          align=F, digits=2,
          notes = 'Cluster-robust SEs at the municipal level.',
          notes.append = T, header=FALSE)
```

We find that the intervention was sufficient to improve the school completion rate. The TDP application increases in 6.85 percent the chance that a school construction is reported as finished by the end of the intervention. This is in line with our theoretical expectations, but as we discuss below, the finding is not consistent. All other coefficients are not statistically significant at conventional levels. On the one hand, the results indicate that our placebo outcome, the percentage of the invested executed before the intervention, indeed behaves as predicted. On the other hand, we expected the four remaining outcomes to increase after the introduction of the app. The literature on bottom-up accountability argues that more information about the shortcomings of public service provision will mobilise citizens who, in turn, would pressure state agents to produce better social policies. Our results do not lend support to that hypothesis.

We observe a similar pattern in the second intervention. Table \ref{tab:results2} shows that the treatment does not have a statistically significant effect on any our outcomes of interest, including the placebo. This provides evidence that our previous findings are not a statistical fluke, and questions the robustness of the effect of the TDP app on school completion. Nevertheless, in our additional tests we find that the app improved the difference between the baseline and the end-line reported investment by 157.5 percent. This is a large effect, yet it only reaches statistical significance when we employ inverse probability weights with block randomization and one-to-one nearest neighbourhood matching. The result is in the Supplementary Material.

```{r table3, echo=FALSE, warning=FALSE, message=FALSE}
```

@@

We also note that the effect sizes are notably small in each of all our original models. Therefore, even if the treatment effects were statistically significant, the TDP app would have a reduced impact over the outcomes. This has an important substantive meaning, as it poses a direct challenge to the bottom-up accountability theory, at least for the school construction outcomes we estimated here.

>>>>

---
title:  |
        | Bottom-Up Accountability and Infrastructure Delivery: Mixed Evidence from Brazil
author: 
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
- "Jessica Voigt[^voigt]"
date: \today
thanks: "We thank Fábio Barros, Guilherme Jardim Duarte, Stephen Herzog, Catarina Roman and David Skarbek for their helpful comments. We are specially grateful to the staff of Transparência Brasil for their excellent research assistance. This research was pre-registered on the EGAP pre-registry tool (<https://egap.org/registration/4505>) and approved by the Brazilian Ministry of Education. The authors thank the 2016 Google Social Impact Challenge for financial support and declare there are no conflict of interest. Replication data and code are available at <http://github.com/umbertomignozzetti/tdp-accountability>."
abstract: "Does grassroots participation improve public service provision? We study the effect of a mobile phone application that allows citizens to monitor school construction projects in 1030 Brazilian municipalities. The app provides a platform where users can submit photos of construction sites, consult independent engineers, and contact their representatives about project delays. Our results show the app has a null impact on public spending and school completion times. Additionally, we find that politicians do not react to individual requests. The results suggest it is difficult to motivate bottom-up accountability in new democracies and question the impact of local monitoring on the quality of public services."
abstractspacing: double
keywords: accountability, Brazil, impact evaluation, state capacity, technology
fontsize: 12pt
margin: 
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
---

@@

We use the TDP app to conduct two experimental interventions and test its impact on six outcomes related to school completion rates and complaints to public authorities. Overall, our results show that providing information to citizens has very limited impact on policy outcomes. We find that the TDP app increased the likelihood of school completion by 7 percent in the first experiment, but the effect is null in the second intervention. We also find that the app increased the percentage of the executed budget by 157.5 percent in our second experiment, but only after nearest neighbourhood matching adjustment. No result reported in the first intervention or in the unmatched sample. Importantly, both interventions have no significant impact on any of the other measures of school investment and completion rates we employed in this study.

@@

# Load packages
pkgs <- c('tidyverse', 'dummies', 
          'haven', 'compareGroups', 
          'devtools', 'lubridate', 'brazilmaps',
          'abjutils', 'lfe', 'cobalt', 
          'MatchIt', 'ri2', 'DeclareDesign', 'knitr',
          'DiagrammeR', 'DiagrammeRsvg', 'rsvg', 
          'flextable', 'officer')
invisible(lapply(pkgs, require, character.only = T))

@@

```{r int1, echo=FALSE, warning=FALSE, message=FALSE, results='hide', cache=TRUE}
load('tdp.RData')

# Load packages
pkgs <- c('tidyverse', 'dummies', 
          'haven', 'compareGroups', 
          'devtools', 'lubridate', 'brazilmaps',
          'abjutils', 'lfe', 'cobalt', 
          'MatchIt', 'ri2', 'DeclareDesign', 'knitr',
          'DiagrammeR', 'DiagrammeRsvg', 'rsvg', 
          'flextable', 'officer')
invisible(lapply(pkgs, require, character.only = T))

@@

resInt1 <- calcATEs(outcs, treats, ctrls, fes, cls, dat = impEvalph1)

# Results intervention 1
library(stargazer)
stargazer(resInt1[[2]][[4]],resInt1[[2]][[8]],resInt1[[2]][[12]],
          resInt1[[2]][[16]],resInt1[[2]][[20]],resInt1[[2]][[24]],
          title = '\\label{tab:results1} Impact Evaluation -- Intervention 1', 
          keep='treatTreatment', type = 'latex',
          covariate.labels = 'ATE',
          column.labels = c("Investment & Investment & Delta & Finished & Cancelled & Updated \\\\ & Before & After & Investment & Construction & Construction & Date \\\\"),
          dep.var.labels.include=F, column.sep.width = '-1pt',
          omit.stat=c('adj.rsq', 'ser'), align=F, digits=2,
          notes = 'Cluster-robust SEs at the municipal level.',
          add.lines = list(c("Controls", "X", "X", "X", "X", "X", "X"),
            c("State fixed effects", "X", "X", "X", "X", "X", "X")),
          notes.append = T, header=FALSE)
```

\begin{table}[!htbp] \centering 
  \caption{\label{tab:results1} Impact Evaluation -- Intervention 1} 
  \label{} 
\begin{tabular}{@{\extracolsep{-1pt}}lcccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{6}{c}{\textit{Dependent variable:}} \\ 
\cline{2-7} 
 & Investment & Investment & Delta & Finished & Cancelled & Updated \\ & Before & After & Investment & Construction & Construction & Date \\
\\[-1.8ex] & (1) & (2) & (3) & (4) & (5) & (6)\\ 
\hline \\[-1.8ex] 
 ATE & $-$0.29 & $-$0.97 & $-$0.68 & 0.07$^{***}$ & 0.001 & 0.06 \\ 
  & (2.40) & (2.48) & (1.26) & (0.01) & (0.001) & (0.04) \\ 
  & & & & & & \\ 
\hline \\[-1.8ex] 
Controls & X & X & X & X & X & X \\ 
State fixed effects & X & X & X & X & X & X \\ 
Observations & 2,352 & 2,352 & 2,352 & 2,352 & 2,352 & 2,352 \\ 
R$^{2}$ & 0.04 & 0.05 & 0.08 & 0.03 & 0.01 & 0.08 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{6}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
 & \multicolumn{6}{r}{Cluster-robust SEs at the municipal level.} \\ 
\end{tabular} 
\end{table}

We find that the intervention was sufficient to improve the school completion rate. The TDP application increases in 7 percent the chance that a school construction is reported as finished by the end of the intervention. This is in line with our theoretical expectations, but as we discuss below, the finding is not consistent. All other coefficients are not statistically significant at conventional levels. On the one hand, the results indicate that our placebo outcome, the percentage of the invested executed before the intervention, indeed behaves as predicted. On the other hand, we expected the four remaining outcomes to increase after the introduction of the app. The literature on bottom-up accountability argues that more information about the shortcomings of public service provision will mobilise citizens who, in turn, would pressure state agents to produce better social policies. Our results do not lend support to that hypothesis.

```{r int2, echo=FALSE, warning=FALSE, message=FALSE, results='hide', cache=TRUE}
load('tdp.RData')

# Load packages
pkgs <- c('tidyverse', 'dummies', 
          'haven', 'compareGroups', 
          'devtools', 'lubridate', 'brazilmaps',
          'abjutils', 'lfe', 'cobalt', 
          'MatchIt', 'ri2', 'DeclareDesign', 'knitr',
          'DiagrammeR', 'DiagrammeRsvg', 'rsvg', 
          'flextable', 'officer')
invisible(lapply(pkgs, require, character.only = T))

outcs <- c('percExecBegin', 'percExecEnd',
           'deltaPercExec', 'isFinished',
           'isCancelled', 'updDate')
treats <- c('treat')
ctrls <- c('logPop2015', 'ideb_ai_2015','ideb_af_2015',
           'log_poorFam2010', 'log_totTransf2016')
fes <- c('state')
cls <- c('municipality')

resInt2 <- calcATEs(outcs, 'treatApp', ctrls, fes, cls, dat = impEvalph2)

# Results intervention 2
library(stargazer)
stargazer(resInt2[[2]][[4]],resInt2[[2]][[8]],resInt2[[2]][[12]],
          resInt2[[2]][[16]],resInt2[[2]][[20]],resInt2[[2]][[24]],
          title = '\\label{tab:results2} Impact Evaluation -- Intervention 2',
          dep.var.labels.include=F, 
          keep='treatAppTreatment', type = 'latex',
          covariate.labels = 'ATE',
          column.labels = c("Investment & Investment & Delta & Finished & Cancelled & Updated \\\\ & Before & After & Investment & Construction & Construction & Date \\\\"),
          column.sep.width = '-1pt',
          omit.stat=c('adj.rsq', 'ser'), align=F, digits=2,
          notes = 'Cluster-robust SEs at the municipal level.',
          add.lines = list(c("Controls", "X", "X", "X", "X", "X", "X"),
            c("State fixed effects", "X", "X", "X", "X", "X", "X")),
          notes.append = T, header=FALSE)
```

\begin{table}[!htbp] \centering 
  \caption{\label{tab:results2} Impact Evaluation -- Intervention 2} 
  \label{} 
\begin{tabular}{@{\extracolsep{-1pt}}lcccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{6}{c}{\textit{Dependent variable:}} \\ 
\cline{2-7} 
 & Investment & Investment & Delta & Finished & Cancelled & Updated \\ & Before & After & Investment & Construction & Construction & Date \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4) & (5) & (6)\\ 
\hline \\[-1.8ex] 
 ATE & $-$0.60 & 0.05 & 0.65 & 0.003 & 0.002 & 0.00 \\ 
  & (1.63) & (1.70) & (0.62) & (0.02) & (0.003) & (0.00) \\ 
  & & & & & & \\ 
\hline \\[-1.8ex] 
Controls & X & X & X & X & X & X \\ 
State fixed effects & X & X & X & X & X & X \\ 
Observations & 2,304 & 2,304 & 2,304 & 2,304 & 2,304 & 2,304 \\ 
R$^{2}$ & 0.06 & 0.08 & 0.05 & 0.03 & 0.02 &  \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{6}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
 & \multicolumn{6}{r}{Cluster-robust SEs at the municipal level.} \\ 
\end{tabular} 
\end{table}

We observe a similar pattern in the second intervention. Table \ref{tab:results2} shows that the treatment does not have a statistically significant effect on any our outcomes of interest, including the placebo. This provides evidence that our previous findings are not a statistical fluke, and questions the robustness of the effect of the TDP app on school completion. Nevertheless, in our additional tests we find that the app improved the difference between the baseline and the end-line reported investment by 157.5 percent. This is a large effect, yet it only reaches statistical significance when we employ inverse probability weights with block randomisation and one-to-one nearest neighbourhood matching.

@@

We also note that the effect sizes are small in each of our original models. Therefore, even if the treatment effects were statistically significant, the TDP app would have a reduced impact over the outcomes. This has an important substantive meaning, as it poses a direct challenge to the bottom-up accountability theory, at least for the school construction outcomes we estimated here.
